Noisy Work: A Review of The Uncertainty Visualisation Literature

### 2024-03-19 paper.qmd
removed abstract

---
title: "Noisy Work: A Review of The Uncertainty Visualisation Literature"
author: Harriet Mason
bibliography: references.bib
date: last-modified
toc: true
number-sections: true
format: pdf
editor_options: 
  chunk_output_type: console
abstract: In reviewing the literature of uncertainty visualisation
---
# Current Todo List
This is a list of tasks that need to be completed before I consider the paper complete. Please keep this in mind when reading.

@@

# Are you measuing uncertainty as noise or signal?
Evaluation experiments While uncertainty information *is* often used to improve decisions, it seems the main motivation in its communication is to offer transparency to the viewer about a statistical analysis. That is, it acts as a form of statistical "hedging" for signals found in an analysis. Interviews with experts in statistic back up this primary motivation, as ignoring uncertainty information of often expressed as being similar to fraud or lying and the goal of "improving decisions" is only seen as a secondary outcome of an appropriately hedged signal [@Hullman2020a; @Manski2020]. This concept of uncertainty means that the work on communicating risk does not cover the field of uncertainty communication as a whole. It covers the specific case where the signal itself *is* uncertain, but this work does not naturally extend to the goal of hedging a signal (i.e. when we want to communicate uncertainty as noise).

@@

@@

### Types of Uncertainty Visualisation Papers

The uncertainty visualisation evaluation study asks participants to perform some value abstraction either explicitly by asking participants to report values or vaguely by asking participants to find regions of least certainty, rank plots, report their confidence, and/or make decisions according to uncertainty information [@Hullman201]. Experimenters then compare these results based on the participants accuracy, preference, confidence, and/or speed [@Hullman2016].

## Uncertainty versus Risk
Uncertainty is noise. This may be an obvious statement, but what is not obvious is that the existing literature sees noise as a signal. When we directly ask "What is the varinance of this variable" the variance *becomes* the signal. Noise and signal are not an inherrent property of variance, or meta-data or the aspects of an analysis that are usually secondary considerations, it is a property of the data that *promotes* a message we are trying to infer and the data that *supresses* it. The replication problem in uncertainty visualisation, where experiments are simply repeating established visualisation work, comes from the misunderstandings established in previous sections. Uncertainty is poorly defined, therefore authors do not understand it is not an inherrent property of the data and is actually related to a particular inference question, this incorrect view that uncertainty is always variance or probabilities leads to the class of "uncertainty visualisations" that depict different mathematical objects or visual features with little in the sense of consistent rules, this complicated class has caused visualisation authors to miss that they are simply repeating the established evaluation results from normal information visualisations. This situation is a bit of a mess. It also highlights a unique and facinating problem faced by uncertainty visualisation. If asking direct questions about uncertainty causes us to treat it as a signal, how do we evaluate uncertainty as *noise*?

> Edit note: this section needs a huge rewrite. The idea of this section blends way too much with the measurement discussions in the "noise and signal" section. I have already taken paragraphs from here and just pasted it in the noise section so it might not be coherrent. Additionaly this section goes on SEVERAL long rambles that realistically should have been a sentance or not included at all.

@@

## Decision Making
Decision making questions that typically involve uncertainty discuss the concept of "risk-aversion" as though it is a mistake. The irony of an uncertainty visualisation pper expressing the importance of uncertainty visualisation, but setting up a ground truth or optimum strategy in which participants *should* ignore the uncertainty information is not lose on us. This perspective does not make sense because it implies participants should make decisions entirely based on the average. A possible soltion for this is to provide a utility framework for a particular experiment [@Hullman2016], however it is unclear how easily participants could adopt a utility framework that is different to their own. @Cheong2016 introduced a very simple optimum strategy into their experiment on bushfire uncertainty visualisations, where participants should choose to evacuate a house as soon as its likelihood of catching fire is 50% or higher. Interestingly, despite this very obvious and simple tactic to maximize payout from the experiment, it seems like many participants did not adopt it, instead acting much more warily as though they were considering whether or not they would *actually* evacuate in the event of a fire. These papers also very rarely consider what the optimum decision would be in a given scenario, and a difficulty with anticipating the best strategy is conflated with the failings of the uncertainty visualisation. Ultimately, we agree with @Hullman2016 who suggested that what determines an approripate ground truth is largely a philosophical exercise.

## Trust
Authors also occasionally use trust as a proxy for uncertainty. Similar to measuring uncertainty through decision quality, this method also has its own series of issues. If the purpose of displaying uncertainty information is to appropriately hedge a signal with noise, then it should be assumed that trust is only related to uncertainty communication through increased transparency and honesty. It does seem to be the case that uncertainty communication increases trust because it is a proxy for transparency *(add citations)*. Unfortunately a large proportion of the literature discusses trust as something that is directly related to uncertainty visualisation rather than as an observable by product. In viewing uncertainty communication as directly related to trust, not related through the proxy of transparency, several unobserved variables are conflated. The amount of uncertainty that is present in the information, whether or not the information is trustworthy or makes sense, prior beliefs of the participants, and the trustworthiness of the source are a few examples of variables that are conflated when authors directly consider trust to be of direct interest. Considering trust, and not transparency, as the metric of importance in uncertainty communication leads to a questionable subtext that argues against transparency. @Hullman2020a found that author simultaneously argued that failing to visualise uncertainty was akin to fraud, but also many avoided uncertainty visualisation because they didn't want their work to come across as "untrustworthy".Being concerned about audiences perception of trust, without first establishing if what we are communicating is *trustworthy*, leads to the implication that details that result in information not being considered trustworthy should be avoided. Authors that imply that a decrease in trust or an increase in ambiguity or risk aversion are metrics of importance in of themselves do not understand the purpose of uncertainty communication as methods to increase transparency. Science communication should be primarily concerned with accuracy, setting trust and risk-aversion as the variables of interest implicitly encourages statisticians to set trust and risk-aversion as the primary goals of communication. The issue of trust being divorced from trustworthiness has been commented on by other authors [@ONeill2018], however the issue still persists in the uncertainty visualisation literature [@Zhao2023].

@@

## Other attemps to capture uncertainty
There are other studies that try to include uncertainty information without directly asking for it but do not go down the route of asking for a decision or levels of trust. These papers often try to ask a question that should utilise both the uncertainty and signal in the response. This is an interesting approach, but these papers are often still looking for a "signal" which is specified in the form of a "ground truth" that is used to evaluate the plot. Unfortunately this usually comes in the form of slightly cryptic questions that create a large amount of noise on the interpretation side [@Hullman2016].

@@

## Uncertainty and hypothesis testing
It is unclear how participants are supposed to incorperate uncertainty information into their responses when uncertainty visualisation authors themselves are unsure. Other authors interpret this difficulty as a combination of the uncertainty and task complexity [@Kinkeldey2014], and not as a general confusion. All these examples sever to illustrate that *we dont know how to measure uncertainty* and if we don't know how to measure it, *we cannot evaluate the quality of a visualisation*. It could be argued that a well done uncertainty visualisations should have an imperceptible signal unless the signal would be identified with a hypothesis, but this idea also has a miriad of issues that come with it.  There is already a series of issues with the reject/do not reject concepts in hypothesis testing, so it is already questionable to set the results of a hypothesis test as the ground truth in an uncertainty visualisation experiment, however these may be additional problems that go beyond the binary nature of the issue. @Patrick2023 compared people ability to recognise patterns in a residual plot to typical statistical tests and found human viewers looking at a plot were less sensitive than the typical residual tests. This increased sensisitivity could be the result of another aspect of uncertainty that is ignored. Statstical tests are typically built upon a series of assumptions, and it is difficult to identify *which* assumption failure caused the data to fail the statistical tests. Lineup plots allow us to see *how* our data is different to the assumed distribution, and better understand *why* our data may have failled a test ^[Did Patricks paper compare statistical tests to lineup protocols in the event an earlier assumption of the tests, such as, independent observation? I think he did check some. I wonder if you can quantify the information difference you get from a rejected statistical test vs a rejected line plot]. These experiments utilised the lineup protocol which has been suggested as a method to check if perceived patterns are real or merely the result of chance [@Buja2009; @Wickham2010; @Chowdhury]. This concept bears similarity to the goal of uncertainty visualisation, but it is not quite the same. @fig-hypyvs shows the conceptual difference between the lineup protocol and uncertainty visualisation. A lineup protocol displays the uncertainty about the null and identifies if the true data plot is identifiable (and therefore significiantly difference) while an uncertainty plot displays the variance of the estimated value and assesses if the null of "no signal" is within this plot (e.g. if the error bars overlap with zero). ^[Does this flip have mathematical implications in how the plots should be assessed?] This implies a visible signal in an uncertainty plot should indicate some divergence from the null.

@@

## Suggestions for measuring uncertainty

>>>>

---
title: "Noisy Work: A Review of The Uncertainty Visualisation Literature"
author: Harriet Mason
bibliography: references.bib
date: last-modified
toc: true
number-sections: true
format: pdf
editor_options: 
  chunk_output_type: console
---
# Current Todo List
This is a list of tasks that need to be completed before I consider the paper complete. Please keep this in mind when reading.

@@

# Are you measuing uncertainty as noise or signal?
Most evaluation experiments fail to capture uncertainty as we want to measure it, as noise. Some authors may read this and think that obviously, we dont use

Uncertainty acts as a form of statistical "hedging" for signals found in an analysis. Interviews with experts in statistic back up this primary motivation, as ignoring uncertainty information of often expressed as being similar to fraud or lying and the goal of "improving decisions" is only seen as a secondary outcome of an appropriately hedged signal [@Hullman2020a; @Manski2020].

While uncertainty information *is* often used to improve decisions, it seems the main motivation in its communication is to offer transparency to the viewer about a statistical analysis.

This concept of uncertainty means that the work on communicating risk does not cover the field of uncertainty communication as a whole. It covers the specific case where the signal itself *is* uncertain, but this work does not naturally extend to the goal of hedging a signal (i.e. when we want to communicate uncertainty as noise).

@@

Uncertainty is noise. This may be an obvious statement, but what is not obvious is that the existing literature sees noise as a signal. When we directly ask "What is the varinance of this variable" the variance *becomes* the signal. Noise and signal are not an inherrent property of variance, or meta-data or the aspects of an analysis that are usually secondary considerations, it is a property of the data that *promotes* a message we are trying to infer and the data that *supresses* it. The replication problem in uncertainty visualisation, where experiments are simply repeating established visualisation work, comes from the misunderstandings established in previous sections. Uncertainty is poorly defined, therefore authors do not understand it is not an inherrent property of the data and is actually related to a particular inference question, this incorrect view that uncertainty is always variance or probabilities leads to the class of "uncertainty visualisations" that depict different mathematical objects or visual features with little in the sense of consistent rules, this complicated class has caused visualisation authors to miss that they are simply repeating the established evaluation results from normal information visualisations. This situation is a bit of a mess. It also highlights a unique and facinating problem faced by uncertainty visualisation. If asking direct questions about uncertainty causes us to treat it as a signal, how do we evaluate uncertainty as *noise*?

@@

@@

### Decision making
Decision making questions that typically involve uncertainty discuss the concept of "risk-aversion" as though it is a mistake. The irony of an uncertainty visualisation pper expressing the importance of uncertainty visualisation, but setting up a ground truth or optimum strategy in which participants *should* ignore the uncertainty information is not lose on us. This perspective does not make sense because it implies participants should make decisions entirely based on the average. A possible soltion for this is to provide a utility framework for a particular experiment [@Hullman2016], however it is unclear how easily participants could adopt a utility framework that is different to their own. @Cheong2016 introduced a very simple optimum strategy into their experiment on bushfire uncertainty visualisations, where participants should choose to evacuate a house as soon as its likelihood of catching fire is 50% or higher. Interestingly, despite this very obvious and simple tactic to maximize payout from the experiment, it seems like many participants did not adopt it, instead acting much more warily as though they were considering whether or not they would *actually* evacuate in the event of a fire. These papers also very rarely consider what the optimum decision would be in a given scenario, and a difficulty with anticipating the best strategy is conflated with the failings of the uncertainty visualisation. Ultimately, we agree with @Hullman2016 who suggested that what determines an approripate ground truth is largely a philosophical exercise.

### Trust and confidence
Authors also occasionally use trust as a proxy for uncertainty. Similar to measuring uncertainty through decision quality, this method also has its own series of issues. If the purpose of displaying uncertainty information is to appropriately hedge a signal with noise, then it should be assumed that trust is only related to uncertainty communication through increased transparency and honesty. It does seem to be the case that uncertainty communication increases trust because it is a proxy for transparency *(add citations)*. Unfortunately a large proportion of the literature discusses trust as something that is directly related to uncertainty visualisation rather than as an observable by product. In viewing uncertainty communication as directly related to trust, not related through the proxy of transparency, several unobserved variables are conflated. The amount of uncertainty that is present in the information, whether or not the information is trustworthy or makes sense, prior beliefs of the participants, and the trustworthiness of the source are a few examples of variables that are conflated when authors directly consider trust to be of direct interest. Considering trust, and not transparency, as the metric of importance in uncertainty communication leads to a questionable subtext that argues against transparency. @Hullman2020a found that author simultaneously argued that failing to visualise uncertainty was akin to fraud, but also many avoided uncertainty visualisation because they didn't want their work to come across as "untrustworthy".Being concerned about audiences perception of trust, without first establishing if what we are communicating is *trustworthy*, leads to the implication that details that result in information not being considered trustworthy should be avoided. Authors that imply that a decrease in trust or an increase in ambiguity or risk aversion are metrics of importance in of themselves do not understand the purpose of uncertainty communication as methods to increase transparency. Science communication should be primarily concerned with accuracy, setting trust and risk-aversion as the variables of interest implicitly encourages statisticians to set trust and risk-aversion as the primary goals of communication. The issue of trust being divorced from trustworthiness has been commented on by other authors [@ONeill2018], however the issue still persists in the uncertainty visualisation literature [@Zhao2023].

@@

### Other attemps to capture uncertainty
There are other studies that try to include uncertainty information without directly asking for it but do not go down the route of asking for a decision or levels of trust. These papers often try to ask a question that should utilise both the uncertainty and signal in the response. This is an interesting approach, but these papers are often still looking for a "signal" which is specified in the form of a "ground truth" that is used to evaluate the plot. Unfortunately this usually comes in the form of slightly cryptic questions that create a large amount of noise on the interpretation side [@Hullman2016].

@@

## Hypothesis testing versus uncertainty
It is unclear how participants are supposed to incorperate uncertainty information into their responses when uncertainty visualisation authors themselves are unsure. Other authors interpret this difficulty as a combination of the uncertainty and task complexity [@Kinkeldey2014], and not as a general confusion. All these examples sever to illustrate that *we dont know how to measure uncertainty* and if we don't know how to measure it, *we cannot evaluate the quality of a visualisation*. It could be argued that a well done uncertainty visualisations should have an imperceptible signal unless the signal would be identified with a hypothesis, but this idea also has a miriad of issues that come with it.  There is already a series of issues with the reject/do not reject concepts in hypothesis testing, so it is already questionable to set the results of a hypothesis test as the ground truth in an uncertainty visualisation experiment, however these may be additional problems that go beyond the binary nature of the issue. @Patrick2023 compared people ability to recognise patterns in a residual plot to typical statistical tests and found human viewers looking at a plot were less sensitive than the typical residual tests. This increased sensisitivity could be the result of another aspect of uncertainty that is ignored. Statstical tests are typically built upon a series of assumptions, and it is difficult to identify *which* assumption failure caused the data to fail the statistical tests. Lineup plots allow us to see *how* our data is different to the assumed distribution, and better understand *why* our data may have failled a test ^[Did Patricks paper compare statistical tests to lineup protocols in the event an earlier assumption of the tests, such as, independent observation? I think he did check some. I wonder if you can quantify the information difference you get from a rejected statistical test vs a rejected line plot]. These experiments utilised the lineup protocol which has been suggested as a method to check if perceived patterns are real or merely the result of chance [@Buja2009; @Wickham2010; @Chowdhury]. This concept bears similarity to the goal of uncertainty visualisation, but it is not quite the same. @fig-hypyvs shows the conceptual difference between the lineup protocol and uncertainty visualisation. A lineup protocol displays the uncertainty about the null and identifies if the true data plot is identifiable (and therefore significiantly difference) while an uncertainty plot displays the variance of the estimated value and assesses if the null of "no signal" is within this plot (e.g. if the error bars overlap with zero). ^[Does this flip have mathematical implications in how the plots should be assessed?] This implies a visible signal in an uncertainty plot should indicate some divergence from the null.

@@
